Estimating duration using survival analysis

### 2024-02-19 ATACCC.tex
ch5 proof

Two major approaches have been taken to model SARS-CoV-2 viral load trajectories, both based on well-well-established methods within biostatisics.
The first approach is an ODE-based approach uses parameters representing biological properties of the virus and host~\autocite[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
The model here is an application of a standard model for viral dynamics, used for many diseases~\autocite{perelsonMechanistic,hernandez-vargasInhost}.
Such models are rarely analytically-tractable and the system's evolution is computed using numerical methods.
% Both of these approaches have been used with other viruses and biomarkers.
The second approach, and the focus of this chapter, is using a piecewise linear model for the biomaker trajectory and random effects to allow for within-individual correlation.

>>>>

Two major approaches have been taken to model SARS-CoV-2 viral load trajectories, both based on well-well-established methods within biostatisics.
The first approach is an ODE-based approach uses parameters representing biological properties of the virus and host~\autocite[e.g.][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
The model here is an application of a standard model for viral dynamics, used for many diseases~\autocite{perelsonMechanistic,hernandez-vargasInhost}.
Such models are rarely analytically-tractable and the system's evolution is computed using numerical methods.
% Both of these approaches have been used with other viruses and biomarkers.
The second approach, and the focus of this chapter, is using a piecewise linear model for the biomaker trajectory and random effects to allow for within-individual correlation.

### 2024-02-19 cis-perfect-testing.tex
ch5 proof

\section{Definitions}
\todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
\begin{itemize}
\item
  We assume that, for all episodes $i$, $D_i$ is iid and independent
  of the time of the infection. Define the survival function
  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
  where $\theta$ are the parameters controlling the survival
  distribution (the discussion in this document is valid regardless of
  the model specified for $S_\theta$ and hence we consider $\theta$
  as an arbitrary vector of parameters).
\item
  The beginning of episode $i$ is known to occur in the interval
  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
  in $[l_i^{(e)}, r_i^{(e)}]$.
\end{itemize}

@@

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estimates of the mean are sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, model assumptions drive the size of the estimated tails.
In this chapter, I develop a method to estimate the duration distribution using CIS data.
The CIS sampling is less frequent, but with a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
I use a simulation study (see \cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (described in \cref{perf-test:sec:parameters-priors}).
These include a prior to combine the information in the CIS with the results of \cref{E-ATACCC}.
Reasonable priors successfully recover the true distribution.
However, the method must be able to account for false negatives, discussed in \cref{E-imperf-test}.

@@

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections.
This is because the interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days, see\todo{ref ATACCC chapter}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

\subsection{Double interval censoring}

@@

In the rest of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC}.
This choice is motivated by computational efficiency and limitations of the CIS's study design.
The biomarker approach is computational challenging because the number of random effects scales with the number of detected infection episodes.
There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
Therefore, the biomarker approach would be computationally challenging.
The computational challenge is compounded by the data being stored in a TRE, which has limited computational power and software.
The issue with random effects is made worse by the CIS's study design.
Infrequent testing means that most detected episodes have only a single positive test.
Hence, there is no information on the viral load trajectory and the random effects are unidentifiable.

@@

Undetected events where there is no information on these events is known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
There is a subtlety that makes the CIS different from standard left truncation.
In CIS, individuals in which undetected infections occurred are enrolled in the study.
Therefore, some data has been collected about them.
In standard left truncation, nothing is known about the missed events.
I incorporate the additional information into the analysis.
I refer to these events as ``undetected'' to emphasise this difference.

@@

Few studies look at the combination of both double interval censoring and undetected (or truncated) events
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}.
The frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs.
They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.\todo{this is a bit unclear, try and rephrase?}

@@

I use Bayesian inference for two reasons.
First, frequentist methods for estimating the parameter's variance and covariance are poorly developed.
Commonly used methods are based on the Fisher information matrix or resampling.
Neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard.
I explore priors in section 1.4.
These prior structures overcome issues caused by the undetected episodes ~\autocite{caoBias}.

Bayesian methods been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling}, augmenting the data with the unobserved times of infection.
The augmentation creates a parameter per detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible here.
I develop a novel Bayesian framework without explicit data augmentation.
This method scales to the large number of episodes present in CIS.

@@

Integrating over the number of missed episodes is analytically tractable.
This makes inference tractable within modern, high-performance Bayesian software such as Stan.
Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
Let $\Ncis$ be the number of individuals in the cohort and $n_\sched$ be the number of test schedules.
%Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
Let $i = 1, \dots \ntot$ index infection episodes.

@@

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
  This space is empty for $i \notin \set{D}$.
  Otherwise, define $l_i^{(b)}$ and $r_i^{(b)}$ as the first and last days $b_i$ could have occurred.
  $l_i^{(b)}$ is the day after the last negative test prior to the first positive test in the episode.
  $r_i^{(b)}$ is the day of the first positive test.
  Similarly, define $l_i^{(e)}$ and $r_i^{(e)}$ as the first and last days $e_i$ could have occurred.
  If $i \in \set{D}$ then $n_{ia} =1$ admissible episodes occurred in individuals with test schedule $\sched_{i}$, otherwise $n_{ia} = 0$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
  \label{perf-test:def:admissible}
\item
  Undetected episodes, $\Omega_i^C$, which have an infection and recovery time such that they would not have tested positive.
  An unknown number, $n_{iu}$ of these observed in individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is undetected with probability $p_{iu} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
\item
  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
  detected.
  This is all remaining episodes not in the previous sets.
  $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in these individuals, it is inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
\end{itemize}

The detected region, $\Omega_i = \alpha_i \cup \beta_i$, for test schedule $\sched_i$ is the region where we could have observed episodes. 
The notation is for consistency with \textcite{heiseyModelling}.
$\Omega_i$ is the complement of $\Omega^C_i$.

@@

These three classes span all possible episodes and are mutually exclusive.
Therefore, these definitions are consistent with the previous definition of $\ntot$.
That is, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\ntot} n_{iu} + 0$.
Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.

@@

In forming the probability $p_{ia}$, the data is used.
This is not quite correct because forming a likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own class is possible.
Then, the counts for all the classes that previously contained undetected episodes would be zero.
The resulting likelihood would be the same.
Therefore, I present this version as it is clearer.

The posterior we are interested in is $p(\theta \mid \na)$.
The posterior density can be written as follows.
\begin{align}
&p(\theta \mid \na) \\
&\propto p(\theta) p(\na \mid \theta) \\
&= p(\theta) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) p(\na \mid \theta, \ntot) \\
&\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=\ndet}^{\ndet} p_{ia} \right) \\
&= p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
\end{align}
The rest of this section derives expressions for each $p_{ia}$, $p_{u}$ and the sum.

Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
I assume that $\theta$, which are only the parameters of the survival distribution, is a priori independent of $\ntot$, the number of infections that occurred in the cohort.
Therefore, $p(\ntot \mid \theta) = p(\ntot)$.
Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

@@

$\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
Its posterior can be reconstructed using a posterior sample of $\theta$.
For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
The full conditional of $\ntot$ is given by:
\begin{align}
p(n_\text{tot} \mid \ncis, \theta)
&\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
&\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
&\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
&= \frac{\Gamma(r + \ntot)}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot}  \\
&= \frac{\Gamma((r + \ndet) + (\ntot- \ndet))}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot-\ndet}.
\end{align}
Comparing this expression to the pmf of a negative binomial, we find that $\nnodet = \ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
%The mean of this distribution is $(r+\ndet)\mu \pnodet/(r+\mu(1-\pnodet))$.

Next, I derive $p_{ia}$.
By definition (see~\cpageref{perf-test:def:admissible}), we have:
\todo{Make sure "detectable" is defined: It's really a latent quantity, which could possibly be defined in terms of true viral load. I think "positive probability of testing positive" under the assumption of no false positives is what I will go with, or by reference to a multi-state model type view of infection.}
\begin{align}
p_{ia} &= \prob((b, e) \in \alpha_i \mid \theta) \\
\alpha_i &= \{ (b, e) : l_i^{(b)} \leq b \leq r_i^{(b)} \wedge l_i^{(e)} \leq e \leq r_i^{(e)}\}.
\intertext{Suppressing the conditioning on $\theta$, this gives:}
p_{ia}
&= \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)}, l_i^{(e)} \leq E_i \leq r_i^{(e)} \right) \\
&= \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \\
&=\sum_{b = l_i^{(l)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid B_i = b \right) \prob \left(B_i = b \right) \\
&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} - b + 1 \leq D_i \leq r_i^{(e)} - b + 1 \right) \prob \left(B_i = b \right) &\text{by def of $D_i$} \\
&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 1) \right) \prob \left(B_i = b \right) &\text{by def of $S_\theta$} \\
&\propto \sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 1) \right)
\end{align}
under the assumption of uniform probability of infection time.

>>>>

% \section{Definitions}
% \todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
% \begin{itemize}
% \item
%   We assume that, for all episodes $i$, $D_i$ is iid and independent
%   of the time of the infection. Define the survival function
%   $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
%   where $\theta$ are the parameters controlling the survival
%   distribution (the discussion in this document is valid regardless of
%   the model specified for $S_\theta$ and hence we consider $\theta$
%   as an arbitrary vector of parameters).
% \item
%   The beginning of episode $i$ is known to occur in the interval
%   $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
%   in $[l_i^{(e)}, r_i^{(e)}]$.
% \end{itemize}

@@

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estimates of the mean are sensitive to the tail of the duration distribution.

In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up, enabling estimation of the tail.
Therefore, model assumptions drive the size of the estimated tail.

In this chapter, I develop a method to estimate the duration distribution using CIS data.
The CIS sampling is less frequent, but with a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (\cref{perf-test:sec:problem}).
I use a simulation study (\cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (\cref{perf-test:sec:parameters-priors}).
These include a prior to combine the information in the CIS with the results of \cref{E-ATACCC}.
Reasonable priors successfully recover the true distribution.
However, the method must be able to account for false negatives, discussed in \cref{E-imperf-test}.

\todo[inline]{make sure following paragraph makes sense in context (copy/pasted here)}
In the rest of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC}.
This choice is motivated by computational efficiency and limitations of the CIS's study design.
The biomarker approach is computational challenging because the number of random effects scales with the number of detected infection episodes.
There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
Therefore, the biomarker approach would be computationally challenging.
The computational challenge is compounded by the data being stored in a TRE, which has limited computational power and software.
The issue with random effects is made worse by the CIS's study design.
Infrequent testing means that most detected episodes have only a single positive test.
Hence, there is no information on the viral load trajectory and the random effects are unidentifiable.

@@

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections.
This is because the interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

\subsection{Double interval censoring}
\todo{using double interval censored before its defined}

@@

@@

Undetected events where there is no information on these events is known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
There is a subtlety that makes the CIS different from standard left truncation.
In CIS, individuals in which undetected infections occurred are enrolled in the study.
Therefore, some data has been collected about them.
In standard left truncation, nothing is known about the missed events.
Here, I will incorporate the additional information into the analysis.
I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.

@@

Few studies look at the combination of both double interval censoring and undetected (or truncated) events
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}, yet these frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs.
They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.

@@

I use Bayesian inference for two reasons.
First, frequentist estimators are poorly developed and their properties (\eg consistency and convergence rate) are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard.
I explore priors in section 1.4.
These prior structures overcome issues caused by the undetected episodes ~\autocite{caoBias}.

Bayesian methods have been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling} by augmenting the data with the unobserved times of infection.
The augmentation creates a parameter per detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible within the SRS.
I develop a novel Bayesian framework without explicit data augmentation.
This method scales to the large number of episodes present in CIS.

@@

Integrating over the number of missed episodes is possible analytically.
This makes inference tractable within modern, high-performance Bayesian software such as Stan.
Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
Let $\Ncis$ be the number of individuals in the cohort and $n_\sched$ be the number of test schedules.
%Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
Let $i = 1, \dots \ntot$ index infection episodes.

@@

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
  This space is empty for $i \notin \set{D}$.
  If $i \in \set{D}$ then $n_{ia} =1$ admissible episodes occurred in an individual with test schedule $\sched_{i}$, otherwise $n_{ia} = 0$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
  \label{perf-test:def:admissible}
\item
  Undetected episodes, $\Omega_i^C$, which have an infection and recovery time such that they would not have tested positive.
  An unknown number, $n_{iu}$ of these observed in individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is undetected with probability $p_{iu} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
\item
  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
  detected.
  This is all remaining episodes not in the previous sets.
  $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in these individuals, it is inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
\end{itemize}

% The detected region, $\Omega_i = \alpha_i \cup \beta_i$, for test schedule $\sched_i$ is the region where we could have observed episodes. 
% The notation is for consistency with \textcite{heiseyModelling}.
% $\Omega_i$ is the complement of $\Omega^C_i$.

@@

These three classes span all possible episodes and are mutually exclusive.
Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\ntot} n_{iu} + 0$.
Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.

@@

In forming the probability $p_{ia}$, the data is used.
This is not quite correct because the cell probabilities in the likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own class is possible.
Then, the counts for all the classes that previously contained undetected episodes would be zero.
The resulting likelihood would be the same.
Therefore, I present this version as it is clearer.

The posterior density we are interested in is $p(\theta \mid \na)$.
It can be written as follows.
\begin{align}
&p(\theta \mid \na) \\
&\propto p(\theta) p(\na \mid \theta) \\
&= p(\theta) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) p(\na \mid \theta, \ntot) \\
&\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=\ndet}^{\ndet} p_{ia} \right) \\
&= p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
\end{align}
The rest of this section derives expressions for each $p_{ia}$, $p_{u}$ and the sum.

To start, I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
I assume that $\theta$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
Therefore, $p(\ntot \mid \theta) = p(\ntot)$.
Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

@@

% $\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
% Its posterior can be reconstructed using a posterior sample of $\theta$.
% For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
% The full conditional of $\ntot$ is given by:
% \begin{align}
% p(n_\text{tot} \mid \ncis, \theta)
% &\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
% &\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
% &\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
% &= \frac{\Gamma(r + \ntot)}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot}  \\
% &= \frac{\Gamma((r + \ndet) + (\ntot- \ndet))}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot-\ndet}.
% \end{align}
% Comparing this expression to the pmf of a negative binomial, we find that $\nnodet = \ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
%The mean of this distribution is $(r+\ndet)\mu \pnodet/(r+\mu(1-\pnodet))$.

Next, I derive $p_{ia}$.
By definition (see~\cpageref{perf-test:def:admissible}), we have:
\begin{align}
p_{ia} &= \prob((b, e) \in \alpha_i \mid \theta) \\
\alpha_i &= \{ (b, e) : l_i^{(b)} \leq b \leq r_i^{(b)} \wedge l_i^{(e)} \leq e \leq r_i^{(e)}\}.
\intertext{Suppressing the conditioning on $\theta$, this gives:}
p_{ia}
&= \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)}, l_i^{(e)} \leq E_i \leq r_i^{(e)} \right) \\
&= \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \\
&=\sum_{b = l_i^{(l)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid B_i = b \right) \prob \left(B_i = b \right) \\
&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} - b + 1 \leq D_i \leq r_i^{(e)} - b + 1 \right) \prob \left(B_i = b \right) &\text{by def of $D_i$} \\
&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 1) \right) \prob \left(B_i = b \right) &\text{by def of $S_\theta$} \\
&\propto \sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 1) \right)
\end{align}
under the assumption of uniform probability of infection time.
